---
title: "hw3"
author: "Wangqian Ju, Yudi Zhang"
date: "10/28/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
library(mvtnorm)
library(Matrix)
set.seed(0)
library(plot3D)
library(tidyverse)
```

## q2
```{r}
Vj <- function(tau, sigma, z, nj)
    (sigma^2) * diag(nj) + (tau^2) * z %*% t(z)

Aj <- function(tau, sigma, beta, x, y, z, q, nj, Vj)
  sum(diag(t(z) %*% z %*% ((tau^2)*diag(q) - (tau^4)*t(z) %*% 
                             solve(Vj(tau, sigma, z, nj))%*%z))) + 
  (tau^4)*t(y-x %*% beta) %*% solve(Vj(tau, sigma, z, nj)) %*% z %*%
  t(z) %*% z %*% t(z) %*% solve(Vj(tau, sigma, z, nj)) %*% (y-x %*% beta)

Cj <- function(tau, sigma, beta, x, y, z, q, nj, Vj)
  (tau^4)*t(y-x %*% beta) %*% solve(Vj(tau, sigma, z, nj)) %*%
  z %*% t(z) %*% solve(Vj(tau, sigma, z, nj)) %*% (y-x %*% beta) + 
  sum(diag((tau^2)*diag(q)-(tau^4)*t(z) %*% solve(Vj(tau, sigma, z, nj)) %*% z)) 

Dj <- function(tau, sigma, beta, x, y, z, nj, Vj)
  (tau^2)*z %*% t(z) %*% solve(Vj(tau, sigma, z, nj)) %*% (y-x %*% beta)


em <- function(beta0, tau0, sigma0, X, Y, Z, n, p, q, tol = 1e-5){
  beta_new=beta0
  tau_new=tau0
  sigma_new=sigma0
  N = cumsum(n)
  m <- length(n)
  
  z = list()
  y = list()
  x = list()
  for (j in 1:m) {
    zs=matrix(nrow = n[j],ncol = q)
    xs=matrix(nrow = n[j],ncol = p)
    ys = matrix(nrow = n[j],ncol = 1)
    if(j==1) {
      for (i in 1:n[1])
        for (k in 1:q) 
          zs[i,k]=Z[i,k]
      for (i in 1:n[1]) 
        for (k in 1:p) 
          xs[i,k]=X[i,k]
      for (i in 1:n[1])
          ys[i,1]=Y[i,1]
    }
    if(j>1) {
      for (i in 1:n[j])
        for (k in 1:q)
          zs[i,k]=Z[(N[j-1]+i),(q*(j-1))+k]
      for (i in 1:n[j])
        for (k in 1:p) 
          xs[i,k]=X[(N[j-1]+i),k]
      for (i in 1:n[j])
          ys[i,1]=Y[(N[j-1]+i),1]
    }
    x[[j]] = xs
    z[[j]] = zs
    y[[j]] = ys
  }

  ######E-Step
  A=matrix(nrow = m,ncol = 1)
  C=matrix(nrow = m,ncol = 1)
  for (j in 1:m) {
    A[j] = Aj(tau_new, sigma_new[j], beta_new, x[[j]], y[[j]], z[[j]], q, n[j], Vj)
    C[j] = Cj(tau_new, sigma_new[j], beta_new, x[[j]], y[[j]], z[[j]], q, n[j], Vj)
  }
  siginv = diag(rep((1/sigma_new)^2,n))
  tmp = c()
  for (j in 1:m)
    tmp=c(tmp, Dj(tau_new,sigma_new[j],beta_new,x[[j]],y[[j]],z[[j]], n[j], Vj))
  
  ######M-Step
  beta_hat=solve(t(X)%*%siginv%*%X)%*%t(X)%*%siginv%*%(Y-tmp)
  tau_hat=sqrt(sum(C)/(m*q))
  sigma_hat=sigma_new
  for (j in 1:m)
    sigma_hat[j] = 
    sqrt((A[j]-2*t(y[[j]]-(x[[j]]%*%beta_new)) %*% 
            Dj(tau_new,sigma_new[j],beta_new,x[[j]],y[[j]],z[[j]], n[j], Vj)+
            t(y[[j]]-(x[[j]]%*%beta_new)) %*% (y[[j]]-(x[[j]]%*%beta_new)))/n[j])
  
  while((abs(tau_new-tau_hat) >= tol) || 
        (abs(sigma_new-sigma_hat) >= tol) || (abs(beta_new-beta_hat) >= tol)) {
    beta_new <- beta_hat
    sigma_new <- sigma_hat
    tau_new <- tau_hat
    ######E-Step
    for (j in 1:m) {
      A[j]<-Aj(tau_new, sigma_new[j], 
               beta_new, x[[j]], y[[j]], z[[j]], q, n[j], Vj)
      C[j]<-Cj(tau_new, sigma_new[j], 
               beta_new, x[[j]], y[[j]], z[[j]], q, n[j], Vj)
    }
    tmp = c()
    for (j in 1:m)
        tmp=c(tmp, Dj(tau_new,sigma_new[j],
                      beta_new,x[[j]],y[[j]],z[[j]], n[j], Vj))
    
    siginv=diag(rep((1/sigma_new)^2,n))
    ######M-Step
    beta_hat <- solve(t(X)%*%siginv%*%X)%*%t(X)%*%siginv%*%(Y-tmp)
    for (j in 1:m)
      sigma_hat[j] = sqrt((A[j]-2*t(y[[j]]-(x[[j]]%*%beta_new))
      %*% Dj(tau_new,sigma_new[j],beta_new,x[[j]],y[[j]],z[[j]], n[j], Vj) +
        t(y[[j]]-(x[[j]]%*%beta_new)) %*% (y[[j]]-(x[[j]]%*%beta_new)))/n[j])
    tau_hat=sqrt(sum(C)/(m*q))
  }
  
  par <- list(bate = beta_hat, tau = tau_hat, sigma = sigma_hat)
  return(par)
}
```

```{r}
n <- c(10, 10, 10)
p <- 2
q <- 2
m <- length(n)

X <- matrix(runif(sum(n) * p), nrow = sum(n), ncol = p, byrow = T)
Z <- bdiag(matrix(rnorm(n[1] * q), nrow = n[1], ncol = q, byrow = T),
          matrix(rnorm(n[2] * q), nrow = n[2] , ncol = q, byrow = T),
          matrix(rnorm(n[3] * q), nrow = n[3], ncol = q, byrow = T))
Z <- as.matrix(Z)
beta <- matrix(rnorm(p), nrow = p, ncol = 1, byrow = T)
sigma <- rnorm(m)
tau <- rnorm(1)
b <- matrix(rnorm(m * q, 0, tau^2), nrow = m * q, ncol = 1, byrow = T)
e <- matrix(c(rmvnorm(1, rep(0, n[1]), (sigma[1])^2*diag(n[1])),
            rmvnorm(1, rep(0, n[2]), (sigma[2])^2*diag(n[2])),
            rmvnorm(1, rep(0, n[3]), (sigma[3])^2*diag(n[3]))),
            nrow = sum(n), ncol = 1, byrow = T)
Y <- X %*% beta + Z %*% b + e
beta0 <- matrix(c(1, 1),nrow = p,ncol = 1,byrow = T)
tau0 <- 1
sigma0 <- c(1, 1, 1)
em(beta0, tau0, sigma0, X, Y, Z, n, q = q, p = p)
```

## q3

#### 3

The three selected regions are: 

```{r}
d = oro.nifti::readNIfTI(fname = "new_phantom.nii.gz") 
data = d@.Data
data = drop(data)
r1 = data[40:53,95:114 , ]
r2 = data[81:90, 81:90, ]
r3 = data[100:167, 100:167, ]
```

```{r, echo=FALSE}
image(1:256, 1:256, data[,,1])
rect(40, 95, 53, 114)
rect(81, 81, 90, 90)
rect(100, 100, 167, 167)
text(46, 103, "r1")
text(85.5, 85.5, "r2")
text(130, 130, "r3")
```

The functions for the EM algorithm and all other functions used for q3:

```{r}
logf <- function(x, r, rho, sigma1, sigma2, tau) {
  -1/(2*(1-rho^2)) * (r - tau)^2 *
    (cos(x)^2/sigma1^2 - 
       (2*rho*cos(x)*sin(x))/(sigma1*sigma2) + 
       sin(x)^2/sigma2^2)
}

Qfunc.each <- function(par, par.cur, r, sample.size = 10000, FUNC = logf){
  theta.cond <- arms(sample.size, FUNC, -pi, pi, metropolis = TRUE,
                      arguments = list(
                        r=r, sigma1=par.cur[1], sigma2=par.cur[2], 
                        rho=par.cur[3], tau=par.cur[4]
                      ))
  sigma1 <- par[1]
  sigma2 <- par[2]
  rho <- par[3]
  tau <- par[4]

  intt <-
    cos(theta.cond)^2/sigma1^2 -
    (2*rho*sin(theta.cond)*cos(theta.cond))/(sigma1*sigma2) +
    sin(theta.cond)^2/sigma2^2
  intt <- mean(intt)

  result <-
    -log(sigma1) - log(sigma2) - 0.5*log(1-rho^2) + log(r) -
    1/(2*(1-rho^2)) * (r-tau)^2 * intt

  result
}

Qfunc <- function(par, par.cur, r.vec, sample.size=1000) {
  result <- sapply(r.vec, function(r) {
    Qfunc.each(par, par.cur, r, sample.size)
  })

  return(-sum(result))
}

em_q3 <- function(r, iter.max=40, tol=1e-4, sample.size=1000){
  par.cur <- c(sd(r), sd(r), 0.5, mean(r))

  iter <- 0
  while(TRUE){
    iter <- iter + 1

    par.old <- par.cur
    # obtain the estimates
    opt <- optim(par.cur, Qfunc, par.cur=par.cur, 
                 r.vec=r, sample.size=sample.size,
                 method = "L-BFGS-B", lower = c(0.0001,0.0001,-0.99,0.0001),
                 upper=c(1000, 1000, 0.99, 5000))

    # update the current value
    par.cur <- opt$par

    l2.diff <- sum((par.cur-par.old)^2)

    cat(paste("iter:", iter, "l2.diff:", l2.diff, "Q value:", opt$value, "\n"))

    if(l2.diff < tol){
      cat("less than the tolerance. exit\n")
      break
    } else if (iter > iter.max) {
      cat("reach max iteration. exit\n")
      break
    }
  }

  return(par.cur)

}

variance.est.each <- function(par, r, sample.size = 1000, FUNC = logf){
  sigma1 <- par[1]
  sigma2 <- par[2]
  rho <- par[3]
  tau <- par[4]

  theta.cond <- arms(sample.size, FUNC, -pi, pi, metropolis = TRUE,
                     arguments = list(
                       r=r, sigma1=sigma1, sigma2=sigma2, rho=rho, tau=tau
                     ))

  part.sigma1 <- 
    mean(-2*cos(theta.cond)^2/sigma1^3 + 
           (2*rho*sin(theta.cond)*cos(theta.cond))/(sigma1^2*sigma2))
  part.sigma2 <- 
    mean(-2*sin(theta.cond)^2/sigma2^3 + 
           (2*rho*sin(theta.cond)*cos(theta.cond))/(sigma1*sigma2^2))
  part.rho1 <- 
    mean(cos(theta.cond)^2/sigma1^2 - 
           (2*rho*cos(theta.cond)*sin(theta.cond))/(sigma1*sigma2) +
    sin(theta.cond)^2/sigma2^2)
  part.rho2 <- mean(-2*sin(theta.cond)*cos(theta.cond)/(sigma1*sigma2))
  part.tau <- part.rho1

  dev.sigma1 <- -1/sigma1 - 1/(2*(1-rho^2))*(r-tau)^2*part.sigma1
  dev.sigma2 <- -1/sigma2 - 1/(2*(1-rho^2))*(r-tau)^2*part.sigma2
  dev.rho <- 
    rho/(1-rho^2) + 
    (r-tau)^2 * (-rho/(1-rho^2)^2 * part.rho1 - 1/(2*(1-rho^2)) * part.rho2)
  dev.tau <- (r-tau)/(1-rho^2) * part.tau

  q <- c(dev.sigma1, dev.sigma2, dev.rho, dev.tau)

  I.matrix <- q %*% t(q)

  return(I.matrix)
}

variance.est <- function(par, r.vec, sample.size = 1000, return_inv = TRUE){
  result <- lapply(r.vec, function(r) {
    variance.est.each(par, r, sample.size)
  })

  result <- Reduce("+", result)

  if(return_inv) {
    return(solve(result))
  } else {
    return(result)
  }

}
```

Let $\tau = || \mu ||$, the estimates and variance estimates of $\sigma_1$, $\sigma_2$, $\rho$, and $\tau$ is showing in the following table

```{r, echo=FALSE}
load("q3_all_results.rda")
result.container <- expand.grid(region=1:3, image=1:18) %>% as_tibble()

result.container <- result.container %>% mutate(
  estimates = purrr::map2(region, image, function(j, i) {
    ests[[i]][[j]]
  }),
  variance.matrix = purrr::map2(region, image, function(j, i) {
    ests.var[[i]][[j]]
  }),
  variance = purrr::map(variance.matrix, diag)
)

p1 <- result.container %>% mutate(
  tau.value = purrr::map_dbl(estimates, function(est) est[4]),
  sigma.ratio = purrr::map_dbl(estimates, function(est) est[1] / est[2])
) %>% pivot_longer(cols = c(tau.value, sigma.ratio), names_to = "type", values_to = "value") %>%
  ggplot() +
  geom_line(aes(x=image, y=value, color=factor(region))) +
  facet_wrap(~type, scales = "free")


result.container.all <- result.container %>% unnest(c(estimates, variance))
result.container.all$parameter <- rep(c("sigma1", "sigma2", "rho", "tau"), 18 * 3)

result.container.all <- result.container.all %>% mutate(
  lower = estimates - qnorm(.975) * sqrt(variance),
  upper = estimates + qnorm(.975) * sqrt(variance)
)

r2 <- result.container %>% mutate(
  var.sig1_sig2 = purrr::map_dbl(variance.matrix, function(vv) {
    t(c(1,-1,0,0)) %*% vv %*% c(1,-1,0,0)
  }),
  sig1_sig2 = purrr::map_dbl(estimates, function(est) est[1] - est[2]),
  lower = sig1_sig2 - qnorm(.975) * sqrt(var.sig1_sig2),
  upper = sig1_sig2 + qnorm(.975) * sqrt(var.sig1_sig2)
)

```

```{r, echo=FALSE}
r1 <- result.container.all %>% dplyr::select(image, region, parameter, estimates, variance, lower , upper)

knitr::kable(r1, digits=4)
```

We also plotted $\sigma_1 / \sigma_2$ and $\tau$ for all three regions over the 18 images. It seems that there is some pattern for the normed mean intensity

```{r, echo=FALSE}
p1 + 
  theme_bw()
```

The variance estimates of $\rho$ have been shown in the previous table, and the variance estimates of $\hat{\sigma}_1- \hat{\sigma}_2$ are:

```{r, echo=FALSE}
knitr::kable(r2 %>% dplyr::select(image, region, sig1_sig2, var.sig1_sig2, lower, upper),
             digits = 4)
```

It seems that the corresponding confidence intervals for $\rho$ do not include 0, but those for $\hat{\sigma}_1- \hat{\sigma}_2$ include 0.




